---
title: "Cryptocurrency Web Trends and Market Price"
author: "Alfredo Muñoz Gutierrez"
date: "2022-11-12"
output:
  pdf_document: default
  html_notebook: default
---

## Introduction
On this project I analyzed the relation between the market price of the
top cryptocurrencys and the Web searches for crypto terms. In order to do the 
analysis I got price data of **Bitcoin and Ethereum** from Yahoo Finance website 
and **Google trends** data about keywords related with the crypto ecosystem. 
The main goal of this analysis was find out if there is a positive correlation 
between cryptocurrencys price and Google searches of crypto-related words, 
in case it is true, if an increase in **cryptocurrencys price** is related to a 
**high volume of searches** in the web for crypto terms or if the increase of 
web searches influence the cryptocurrencys price. Other objectives were spot 
the locations with most interest in the crypto terms as well as if there are any 
correlation between Bitcoin and Ethereum price.

### Objetives
#### Answer the following questions:

1. ¿What is the relation between cryptocurrencys market price and Google
trends of crypto terms? ¿Is one variable causing the other to change?

2. ¿Wich locations have most interest in search for crypto terms?

3. ¿Is there a correlation between Bitcoin and Ethereum price?


## Data Collection, Cleaning and Exploration
The first step was to load the librarys `gtrendsR` to get the Google trends data
and `tidyverse` to clean, tidy and plot the data in order to answer the
analysis' questions. The market price data was imported from csv files.
```{r, warning=FALSE}
library(gtrendsR)
library(tidyverse)
setwd("~/R/Codecademy-Capstone Project Exploring Google Trends Data/Data")
#import Bitcoin and Ethereum price data to data frames
btc_data <- read_csv("BTC-USD.csv",show_col_types = FALSE)
eth_data <- read_csv("ETH-USD.csv", show_col_types = FALSE)
```

Then the `btc_data` dataframe was cleaned and the Bitcoin and Ethereum market 
prices were plotted (data from the last 5 years).

```{r}
#Convert date column to date type
btc_data <- btc_data %>% 
  mutate(Date = as.Date(Date,"%d/%m/%Y"))

#Plot Bitcoin price from the last 5 years
ggplot(btc_data) + geom_line(aes(Date,Close), color="#00BA38") +
  labs(title="Bitcoin Market Price",
       caption="Data from Yahoo Finance", x=NULL, y="Market price") +
  scale_y_continuous(label = function(x) paste0('$', x))

```

```{r}
#Plot Ethereum price from the last 5 years
ggplot(eth_data) + geom_line(aes(Date,Close), color="#619CFF") +
  labs(title="Ethereum Market Price",
       caption="Data from Yahoo Finance", x=NULL, y="Market price") +
  scale_y_continuous(label = function(x) paste0('$', x))
```

With the `gtrends` function the popularity over time for the crypto words was 
gotten and after cleaning the data I plotted the results for each keyword.

```{r}
#Get Google trends data about crypto keywords from the last 5 years
crypto_trends <- gtrends(c("cryptocurrency", "blockchain", "NFT"))
crypto_trends_over_time <- crypto_trends$interest_over_time

#Change the "<1" values in "hits" column to 0 and convert to numeric data type
crypto_trends_over_time$hits[crypto_trends_over_time$hits == "<1"] <- "0"
crypto_trends_over_time <- crypto_trends_over_time %>% 
  mutate(hits=as.numeric(hits))

#Plot trends linear graph
ggplot(crypto_trends_over_time, aes(date, hits, color=keyword)) +
  geom_line() +
  scale_color_discrete(name = "Crypto Terms") + 
  labs(title = "Google Trend Data For Crypto Terms",
       x=NULL, y = "Hits (Normalized to be between 0 and 100)")
```
We can see that there was three peaks for "cryptocurrency" searches, first in
early 2018, and the other two in 2021. For "blockchain" the trend is 
similar but is far less popular. "NFT" searches were almost 0 before 2021, 
then it starts to increase until it reaches the top value of 100 in the 
beginning of 2022 (meaning that this was the most searched word in that moment)
but after this peak it starts to suddenly decrease until it reaches around a 
value of 20.

## Crypto Popularity in the World
In order to find what are the places with most interest in search for crypto
terms I use the *interest_by_country* dataframe from the Google trends data
wich has the popularity data for each country. Then I order and filter the 
dataset to get the top 10 countries with most interest in "cryptocurrency"
and "NFT" keywords.

```{r}
#Get top 10 countries with most "cryptocurrency" searches popularity
crypto_trends_countries <- crypto_trends$interest_by_country 
cryptocurrency_countries <- crypto_trends_countries %>% 
  filter(keyword=="cryptocurrency") %>% 
  arrange(-hits) %>% 
  head(10)

#Plot top 10 countries for "cryptocurrency" 
ggplot(cryptocurrency_countries, aes(x=hits, 
      y=forcats::fct_reorder(location, hits))) + geom_col(fill="#00BA38") +
  labs(title="Top 10 countries most Interested in Cryptocurrency Keyword",
       x= "Hits (Normalized to be between 0 and 100)" , y=NULL)

```
Besides being a underdeveloped country **Nigeria** is the place with most
interest in search "cryptocurrency" in Google then this could be a opportunity
for crypto exchanges or another crypto-related business to approach and expand
to this market. 


```{r}
#Get top 10 countries with most "NFT" searches popularity
NFT_countries <- crypto_trends_countries %>% 
  filter(keyword=="NFT") %>% 
  arrange(-hits) %>% 
  head(10)

#Plot 10 countries for "NFT"
ggplot(NFT_countries, aes(x=hits, 
      y=forcats::fct_reorder(location, hits))) + geom_col(fill="#619CFF") +
  labs(title="Top 10 countries most Interested in NFT Keyword",
       x= "Hits (Normalized to be between 0 and 100)" , y=NULL)
```

With 6 of the 10 top countries being from Asia there is no doubt that this is 
the region most interested in NFTs then a  NFT related company that has intentions 
to expand to new markets **China** is the best stake but **Nigeria** is a good
option too due to it is the 4th country most interested in NFTs and the most
interested in cryptocurrencys.

## Price Data vs Trends Data
To be able to answer the main question of the analysis I combine the market 
price and Google trends data adjusting the scale to have a good view of the 
correlation between the variables. Since Bitcoin is the top cryptocurrency in
the market and the most popular I compare its price with the popularity of
"cryptocurrency" keyword and due to Ethereum blockchain is the most used to
buy and sell NFTs I compare the Ethereum price with the "NFT" keyword.

```{r}
#Plot price and trends together for Bitcoin and cryptocurrency 
ggplot() + geom_line(btc_data, mapping = aes(Date, Close, 
  color = "Bitcoin Price")) + geom_line(filter(crypto_trends_over_time, 
  keyword == "cryptocurrency"), mapping = aes(as.Date(date), hits*1200,
  color = "Cryptocurrency Google Trends Hits")) +
  scale_y_continuous(name = "Price", sec.axis = sec_axis(~./1200,
  name="Google Trends Hits")) + 
  labs(title = "Bitcoin Price and Cryptocurrency Google Trends", x=NULL)

```
We can see that in some time periods there is acorrelation between Bitcoin price and 
"cryptocurrency" web searches since when there is a increase in one the other 
increase too but this not happens always. In early 2021 there is a suddenly 
increase in both variables then the market price kept increasing while the 
"cryptocurrency" popularity decrease but when the market price reaches a peak
the web searches suddenly increase reaching a peak at the same time.

```{r}
#Plot price and trends together for Ethereum and NFT 
ggplot() + geom_line(eth_data, mapping = aes(Date, Close, 
  color = "Ethereum Price")) + geom_line(filter(crypto_trends_over_time, 
  keyword == "NFT"), mapping = aes(as.Date(date), hits*50,
  color = "NFT Google Trends Hits")) +
  scale_y_continuous(name = "Price", sec.axis = sec_axis(~./50,
  name="Google Trends Hits")) + 
  labs(title = "Ethereum Price and NFT Google Trends", x=NULL)
```

Similar to the last graph it seems like the Ethereum price increase when the
"NFT" searches does it perhaps a decrease in the web searches does not affect

## Price correlation
Bitcoin and Ethereum seems to be very dependent on the crypto popularity and 
they also look like them prices have a similar trend. I made a scatter plot of
the prices for both cryptocurrencys with a trend line and the calculated linear
regression to see if there is a strong correlation.
```{r warning=FALSE}
#Merge Bitcoin and Ethereum price
crypto_price <- btc_data %>% 
  inner_join(eth_data, by="Date") %>% 
  select(Date,Close.x, Close.y )

#Plot Bitcoin and Ethereum price correlation
ggplot(crypto_price, aes(Close.x,Close.y)) + geom_point() +
  geom_smooth(method= "lm") +
  geom_smooth(se = FALSE, color = "green") +
  labs(title="Bitcoin vs Ethereum Price", 
       x="Bitcoin Price", y="Ethereum Price") +
  scale_y_continuous(label = function(x) paste0('$', x)) +
  scale_x_continuous(label = function(x) paste0("$", x))
```
It looks like there are a positive correlation between the Bitcoin and Ethereum
prices but there some points that does not have the same trend making the trend
line differ from the trend line obtain from the linear regression method.
Possibly if other variables as the date were included in the model better 
results could be achieved.

## Conclusion and Further Steps
In the analysis we explore the effects of crypto terms web popularity in the 
market price of Bitcoin and Ethereum. I can argue that indeed there are a **strong
correlation between this variables** despite they don't follow the same trend
sometimes, **in the long-term if there is a increase in the cryptocurrencys price 
the web searches for crypto terms will increase too**.

However, I cannot say that this statement applies to all the cryptocurrencys or 
that the market price is the only cause for the increase in web popularity for 
crypto terms due to the limitations of the data I worked with. If **more crypto
currencys prices** were included and data about other market assets (like the 
**S&P500**) that reflects better the global market situation a more detalled 
and accurate analysis could be maded.



